### 2011-04-01 Spatial_adaptation/standing-variation.tex
/* Standing variation */

>>>>

\section{Notes on branching process approximation}

The branching process approximation will be good as long as the probability of "interactions" are small,
i.e.\ we can couple the branching process with the actual density--regulated model.
Let $Z_t(x)$ be the numbers of selected individuals in deme $x$.
If we assume that $Z_t$ is smooth enough, 
the numbers of nearby individuals to interact with is approximately $Z_t(x) \sigma^d$,
and the total number of individuals is about $\rho \sigma^d$,
so the rate of interactions is about $(Z_t(x) \sigma^d)^2/(\rho \sigma^d)$.
Now if $Z$ is a spatial branching process beginning from a single individual at $x=0$ at $t=0$, 
and growing at rate $r$, then 
\[ 
    Z_t(x) \approx \exp(rt - \|x\|^2/(2\sigma^2 t))/ (2 \pi \sigma^2)^{d/2} 
        \le e^{rt} / ( \sigma^2 t)^{d/2}
\]
and so we need $t$ such that
\[
    e^{2rt} t^{-d} << \rho \sigma^{3d} .
\]
or equivalently,
\[
    e^{2rt/d} t^{-1} << \rho^{1/d} \sigma^3 .
\]
This will be true if $t$ is of smaller order than the logarithm of the effective population density, $\log \rho \sigma^d$.

### 2011-04-01 standing-patches-paper.tex
/* Standing variation */

For now we ignore geographic hetergeneity, and assume that the species range
is a homogenous, two-dimensional region.  (results for a one-dimensional region also appear)
There are two selective types -- the neutral type, and the mutated type.
We assume that different mutations are distinguishable --
either as selectively equivalent mutations, or by linked variation.
At first, the mutated type is at a seletive disadvantage, for long enough to be at selection-mutation equilibrium,
but at a certain time the selective regime changes, so that the mutated type has a selective advantage and quickly spreads to fixation.
After fixation, alleles are either descended
from families of mutants present as standing variation when the selective regime changed,
or from new mutants arising since that time.

We will suppose that before time $t=0$,
the mutant type has fitness $1-s_d$ relative to the neutral type,
and that after time $t=0$,
the mutant type has fitness $1+s_b$,
for positive numbers $s_d$ and $s_b$.
We assume that diploid fitness is additive, or at lesat that the important dynamics are determined by the haploid fitness.
If the mutation is recessive, XXX more complicated things happen.
More concretely, suppose that the species occupies an area with mean density

>>>>

For now we ignore geographic hetergeneity, and assume that the species range
is a homogenous, two-dimensional region.  (results for a one-dimensional region also appear)
There are two selective types -- the neutral type, and the mutated type.
We assume that different mutations are distinguishable --
either as selectively equivalent mutations, or by linked variation.
At first, the mutated type is at a selective disadvantage, for long enough to be at selection-mutation equilibrium,
but at a certain time the selective regime changes, so that the mutated type has a selective advantage and quickly spreads to fixation.
After fixation, alleles are either descended
from families of mutants present as standing variation when the selective regime changed,
or from new mutants arising since that time.

We will suppose that before time $t=0$,
the mutant type has fitness $1-s_d$ relative to the neutral type,
and that after time $t=0$,
the mutant type has fitness $1+s_b$,
for positive numbers $s_d$ and $s_b$.
We assume that diploid fitness is additive, or at least that the important dynamics are determined by the haploid fitness.
(If the mutation is recessive, XXX more complicated things happen.)

For the other parameters,
suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
and that the XXX dispersal variance is $\sigma^2$.
The species occupies an area with mean density
$\rho$ chromosomes per unit area.
(XXX or do diploid? below is haploid.)

We make use of the commonly used approximation
that neglects spatial structure and competition between relatives,
treating the offspring of a new mutant that appears in an area not already occupied by the mutated type
as a (nonspatial) branching process.
After $t=0$, the offspring of a single mutant form (approximately) a branching process with growth rate $s_b$,
so each new mutant establishes locally with probability $p_s \approx 2s_b$.
Therefore, the total rate at which new mutants appear and establish locally after $t=0$ in areas not already occupied by the mutant type
is approximately $\lambda = 2 \mu \rho s_b$.
Before $t=0$, the mutation is deleterious, so the genetic descendents of each new mutation are doomed to extinction,
but may persist for some time.
Consider a mutation that arose at time $-T<0$, and let $Z_s$ be the number of its descendants at time $s-T$.
At time $t=0$, there are $Z_T$ individuals present with the mutation,
and each has probability $p_s \approx 2s_b$ of establishing, approximately independently.
Therefore, the probability that some descendants of this mutation establish and fix locally is
$1-(1-p_s)^Z_T$, XXX continue here

If $Z_t$ is a branching process with $Z_0 = 1$ and $\E[Z_t] = \exp(-s_d t)$, then

For $t \ge 0$, new mutations appear at rate approximately $\lambda = 4 \mu \rho s_b$.
